---
title: "Female academic trajectories (1967-2015)"
author: "Joanna Tang, Sandro Lallas, Rachem Behm"
date: "11/27/2018"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
# Outline

# 0. Load in packages and read in data
# 1a. Line graph for Total Males, Total Females + LM
# 1b. LM
# 2a. Bar graph, with 1 bar for each subject
# 2b. Chi-squared tests
# 3a. Table for male vs female postdoc positions, male vs female non-postdoc positions
# 3b. Wilcoxon signed-rank + Cliff's Delta (effect size)
# 4. Multivariable LM

```


```{r include = FALSE}

# 0. Load packages and read in data

library(tidyverse)
library(vcdExtra)
library(knitr)
library(kableExtra)
library(RColorBrewer)
library(car)
library(effsize)
library(kableExtra)
library(formattable)
library(stargazer)

faculty_salary <- read_csv("Faculty salary data.csv") %>% rename("Years_since" = "Years Since PhD") %>% rename("Years" = "Years Faculty Service") %>% rename("Rank" = "Faculty Rank")
grad_enrollment <- read_csv("Grad enrollment.csv") %>% rename("Males" = "Total Males") %>% rename("Females" = "Total Females") %>% rename("Full_time_males" = "Full-time-males") %>% rename("Full_time_females" = "Full-time-females") %>% rename("Part_time_males" = "Part-time-males") %>% rename("Part_time_females" = "Part-time-females") %>% rename("Private_total" = "Private-total")
median_salary <- read_csv("Median salary for doctoral recipients.csv") %>% rename("Total_male" = "Total employment male") %>% rename("Total_female" = "Total employment femal") %>% rename("Postdoc_male" = "Postdoc male") %>% rename("Postdoc_female" = "Postdoc female") %>% rename("Field" = "Field of study")
phd_field <- read_csv("PhDs by Field.csv") %>% mutate(Percent = Proportion*100)

```



#INTRODUCTION

#METHODS

##Data Collection

##Statistical Analyses

All statistical analyses and figures were created in R (version 3.4.2).  Data for graduate enrollment (1967-2015) were fitted to linear regression models to analyze changes in graduate enrollment over time.  Chi-squared tests were performed to analyze shifts in female PhD recipients by field (1985, 2000, and 2015). Since our data for post-graduate starting salaries were not normally distributed, we used Wilcoxon signed-rank tests to analyze significant differences between male and female salaries for starting postdoctoral and other employment positions (2015).  Data for faculty salaries were fitted to multiple linear regression models to analyze the effects of faculty sex, rank, years in current position, discipline, and number of years since doctoral degree was earned on faculty salaries.


#RESULTS

##Male and female graduate enrollment (1967-2015)

```{r echo = FALSE, message = FALSE, warning= FALSE}

# 1a. Line graph with 2 lines: Total Males, Total Females + LM for each

grad_enrollment_males <- ggplot(grad_enrollment, aes(x = Year)) +
  geom_point(aes(y = Males), color = "blue") +
    geom_smooth(method = lm, se = TRUE, color = "gray20", aes(y = Males)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = function(x) format (x, big.mark = ",", scientific = FALSE), breaks = seq(200000, 1800000, by = 400000), lim = c(200000, 1800000), expand = c(0,0)) +
  xlab("Year") +
  ylab("Male Graduates Enrolled") +
  ggtitle("Male Graduate School Enrollment (1967-2015)") +
  theme_classic() +
  annotate("text", x = 1980, y = 1500000, label = "y = 9,069x - 17,112,153", family = "Times New Roman", size = 5) +
    annotate("text",  x = 1980, y = 1300000, label = "italic(R)^2 == 0.85", parse = TRUE, family = "Times New Roman", size = 5)

grad_enrollment_males

grad_enrollment_females <- ggplot(grad_enrollment, aes(x = Year)) +
  geom_point(aes(y = Females), color = "red") +
    geom_smooth(method = lm, se = TRUE, color = "gray20", aes(y = Females)) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = function(x) format (x, big.mark = ",", scientific = FALSE), breaks = seq(200000, 1800000, by = 400000), lim = c(200000, 1800000), expand = c(0,0)) +
  xlab("Year") +
  ylab("Female Graduates Enrolled") +
  ggtitle("Female Graduate School Enrollment (1967-2015)") +
  theme_classic() +
  annotate("text", x = 1980, y = 1500000, label = "y = 30,130x - 58,960,000", family = "Times New Roman", size = 5) +
    annotate("text",  x = 1980, y = 1300000, label = "italic(R)^2 == 0.98", parse = TRUE, family = "Times New Roman", size = 5)

grad_enrollment_females



```

```{r include = FALSE}

# 1b. LM

grad_enrollment_males_lm <- lm(Males ~ Year, data = grad_enrollment)
summary(grad_enrollment_males_lm)

grad_enrollment_females_lm <-lm(Females ~ Year, data = grad_enrollment)
summary(grad_enrollment_females_lm)

grad_enrollment_males_pearsons <- cor.test(grad_enrollment$Year, grad_enrollment$Males)
grad_enrollment_males_pearsons

grad_enrollment_females_pearsons <- cor.test(grad_enrollment$Year, grad_enrollment$Females)
grad_enrollment_females_pearsons

```

Total graduate enrollment of both males and females increased from 1967 to 2015 (Fig. 1).  Male enrollment increased from 630,701 in 1967 to 1,221,566 in 2015.  Female enrollment increased from 265,364 in 1967 to 1,718,926 in 2015.  Most notably, up until 1988, total male enrollment exceeded total female enrollment; since 1988, total female enrollment has exceeded total male enrollment. The year significantly predicts male graduate enrollment (Pearson's *r* = .92). The overall model (Male enrollment = 9,069 * Year - 17,112,153) explains a significant amount of variance in male graduate enrollment (F(1, 47) = 276, *p* < 0.001, R^2^ = 0.85). The year significantly predicts female graduate enrollment (Pearson's *r* = .99). The overall model (Female enrollment = 30,126 * Year - 58,955,502) explains a significant amount of variance in female graduate enrollment (F(1, 47) = 2,669, *p* < 0.001, R^2^ = 0.98).




##Shifts in female PhD recipients

```{r echo = FALSE, message = FALSE, warning= FALSE}

# 2a. Bar graph with 1 bar for each subject

female_phd <- phd_field %>% 
  filter(Sex == "Female",
         Field == "Physical sciences and earth sciences" |
         Field == "Engineering" |
         Field == "Education" |
         Field == "Humanities and arts",
         Year == "1985" |
         Year == "2000" |
         Year == "2015")

female_phd_graph <- ggplot(female_phd, aes(x = Year)) +
  geom_col(aes(y = Number, fill = Field), position = "dodge") +
  scale_x_continuous(breaks = seq(1985, 2015, by = 15), expand = c(0, 0)) +
  scale_y_continuous(lim = c(0, 5000), expand = c(0,0)) +
  xlab("Year") +
  ylab("Number of Students") +
  ggtitle("Female PhD Recipients by Field") +
  theme_classic()

  
female_phd_graph

# Alternate graph based on field: female_phd_field_graph <- ggplot(female_phd, aes(x = Field,)) + geom_col(aes(y = Number, fill = as.factor(Year)), position = "dodge")

```

```{r echo = FALSE, message = FALSE, warning= FALSE}
# 2b. Pairwise Chi-squared tests

# Converting female_phd into contingency table

female_phd_contingency <- select(female_phd, "Field", "Number", "Year") %>%
  spread(Year, Number) %>% 
  select(-Field)

rownames(female_phd_contingency) <- c("Education", "Engineering", "Humanities & Arts", "Physical & Earth Sciences")

female_phd_chi <- chisq.test(female_phd_contingency)
#chi2 (6) = 2073.3, p < .001, year affects proportion in each field


female_phd_prop <- round(prop.table(as.matrix(female_phd_contingency), 2), 3)

female_phd_table <- female_phd_prop %>% 
  kable(caption = "Proportions of Female PhD Recipients in each Year") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
  
 female_phd_table

```

The number of female PhD recipients in Education increased from 3,491 in 1985 to 4,179 in 2000, but then decreased to 3,502 in 2015 (Fig. 2).  This is the only field for which the number of recipients declined from 2000 to 2015.

The number of female PhD recipients in Engineering increased from 198 in 1985, to 838 in 2000, to 2,301 in 2015 (Fig. 2). 


The number of female PhD recipients in Humanities & Arts increased from 1,392 in 1985, to 2,672 in 2000, to 2,832 in 2015 (Fig. 2).


The number of female PhD recipients in Physical & Earth Sciences increased from 569 in 1985, to 1,022 in 2000, to 1,988 in 2015 (Fig. 2).

There is a significant effect of year on the proportion of PhD recipients in each field ($\chi^2$(6) = 2073.3, *p* < 0.001).

##Male and female salaries for starting postdoctoral and other employment positions (2015)

```{r echo = FALSE, message = FALSE, warning= FALSE}

# 3a. Boxplots for male vs female postdoc positions, male vs female non-postdoc positions


starting_salary_box <- ggplot(median_salary, aes(x = Male, y = Total_male)) +
  geom_boxplot() +
  geom_boxplot(aes(x = Female, y = Total_female)) +
  scale_y_continuous(breaks = seq(50000, 130000, by = 20000), lim = c(50000, 130000)) +
  xlab("Gender") +
  ylab("Median Salary ($)") +
  ggtitle("Median salary for non-Postdocs employment positions") +
  theme_classic()

starting_salary_box

starting_salary_postdoc_box <- ggplot(median_salary, aes(x = Male, y = Postdoc_female)) +
  geom_boxplot() +
  geom_boxplot(aes(x = Female, y = Postdoc_female)) +
  scale_y_continuous(breaks = seq(40000, 70000, by = 10000), lim = c(40000, 70000)) +
  xlab("Gender") +
  ylab("Median Salary ($)") +
  ggtitle("Median salary for Postdoc employment positions") +
  theme_classic()

starting_salary_postdoc_box


```

```{r include = FALSE}

# 3b. Wilcoxon signed-rank + Cliff's Delta (effect size)

starting_salary_mwu <- wilcox.test(median_salary$Total_male, median_salary$Total_female, paired = TRUE)
starting_salary_mwu #W = 101, p = .0026, medians differ significantly

starting_salary_cliff <- cliff.delta(median_salary$Total_male, median_salary$Total_female)
starting_salary_cliff #delta = .213 (small)


starting_salary_postdoc_mwu <- wilcox.test(median_salary$Postdoc_male, median_salary$Postdoc_female, paired = TRUE)
starting_salary_postdoc_mwu #V = 19.5, p = .8884, medians do not differ significantly

starting_salary_postdoc_cliff <- cliff.delta(median_salary$Postdoc_male, median_salary$Postdoc_female)
starting_salary_postdoc_cliff #delta = .04 (small)

```

The median starting non-postdoc employment positions differ significantly for males (median =  75,167 males) and females (median = 71,750 females)  in our samples (V = 101, *p* = 0.0026, $\alpha$ = .05, Cliff’s delta = .21) (Fig. 3).  Thus, median salary does differ significantly between male and female starting non-postdoc employment position salaries.

Although the median starting postdoc positions for males (median =  48,000) was higher than that for females (median = 45,000) in our samples, there is no significant difference (V = 19.5, *p* = 0.89, $\alpha$ = .05, Cliff’s delta = .04) (Fig. 4).  Thus, median salary does not differ significantly between male and female starting postdoc position salaries.

##Academic salaries for professors in U.S. colleges

```{r include = FALSE}

# 4. Multivariable LM

faculty_salary$Rank <- fct_relevel(faculty_salary$Rank, "AsstProf")

faculty_salary_lm_sat <- lm(Salary ~ Sex + Rank + Years + Discipline + Years_since, data = faculty_salary)

summary(faculty_salary_lm_sat) #negative coefficient doesn't make sense

pairs_df <- select(faculty_salary, Years_since, Years, Salary)
pairs(pairs_df)

plot(faculty_salary_lm_sat)

faculty_salary_lm1 <- lm(Salary ~ Sex + Rank + Discipline + Years, data = faculty_salary)
summary(faculty_salary_lm1) #coefficient still negative

faculty_salary_lm2 <- lm(Salary ~ Sex + Rank + Discipline + Years_since, data = faculty_salary)
summary(faculty_salary_lm2)

# Faculty Salary = 67,884 + 4,349.37(SexMale) + 13,104(RankAssocProf) + 46,033(RankProf) + 13,937(DisciplineB) + 61(Years_since)

aic_lm2 <- AIC(faculty_salary_lm2)
aic_lm2 #9097

faculty_salary_lm3 <- lm(Salary ~ Sex + Rank + Discipline, data = faculty_salary)
summary(faculty_salary_lm3)

aic_lm3 <- AIC(faculty_salary_lm3)
aic_lm3 #9095

faculty_salary_lm4 <- lm(Salary ~ Rank + Discipline, data = faculty_salary)
summary(faculty_salary_lm4)

aic_lm4 <- AIC(faculty_salary_lm4)
aic_lm4 #9094

faculty_salary_lm5 <- lm(Salary ~ Rank + Discipline + Years_since + Sex + Years_since*Discipline, data = faculty_salary)
summary(faculty_salary_lm5)

aic_lm5 <- AIC(faculty_salary_lm5)
aic_lm5 #9098

```


```{r stargazer, echo = FALSE, results = 'asis'}

#Regression table of LMs

stargazer(faculty_salary_lm_sat, faculty_salary_lm2, type = "html")

```
Multiple variables potentially predict faculty salary, including faculty sex, rank, years in current position, discipline, and number of years since doctoral degree was earned.  However, our saturated multivariable linear regression model generated a negative coefficient for years in current position, translating into a *decrease* in salary the longer a faculty member stays in a their current position (Fig. 5).  This does not make sense, and a test for collinearity showed multicollinearity between years in current position and years since doctoral degree was earned.  Eliminating years since doctoral degree in favor of years in current position was earned still generates a negative coefficient for years in current position.  However, eliminating years in current position in favor of years since doctoral degree was earned alleviates this conundrum.

However, we decided to eliminate years since doctoral degree was earned because both are correlated with rank.  This model, including only predictors of sex, rank, and discipline, best predicts faculty salary with the least complexity (and had the lowest AIC value).


#DISCUSSION

##Female graduate enrollment is increasing over time
From 1967 to 2015, female graduate enrollment continually increased.  Total female enrollment increased more than total male enrollment.  Enrollment in public universities increased more than enrollment in private universities.  Full-time female enrollment increased more than full-time male enrollment.  Part-time female enrollment increased more than part-time male enrollment.

##Shifts in female PhD recipients over time
Since female graduate enrollment consistently increased over time, we were curious to explore if female enrollment increased in any particular field of study.

##Starting salaries for female PhD recipients in 2015
Since female graduate enrollment consistently increased over time, and increased more than male graduate enrollment, we were curious to explore if starting salaries for female PhD recipients differed from that of male PhD recipients.

##Academic salaries for US professors (2008-2009)

#CONCLUSION

This report explored the differences in graduate degrees pursued and salary between males and females.

Study findings:
(1) Both male and female graduate degree enrollment increased over time; however, females had a much more pronounced upward enrollment trend, and surpassed total male numbers in 1988.
(2) Females pursuing a PhD in education remained relatively constant over the years. Female enrollment in all other fields, most notably STEM faculties, increased dramatically between 1985 and 2015. 
(3) Wilcoxon signed-rank tests showed a significant difference in male and female salaries for non-postdoctoral jobs, but not for postdoctoral employment positions.
(4) A multiple linear regression model to predict professor salary.

This study should be conducted again in the future to test if the differences between male and female education and salary remain constant or continue to differ. Current trends point towards an equalization between sexes over time. This research will become increasingly important as social progress continues and women keep redefining their role in modern-day society.

#REFERENCES
